Stochastic trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2025-02-25 paper/intro.tex
address review comments, clean up equations, create n_c=2 plot

Spectral distributions of matrices and linear operators reveal important properties in numerous applications across physics, chemistry, engineering, and data science: For example, in electronic structure calculations, eigenvalues represent the energy levels that electrons occupy~\cite{drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure, lin-2017-randomized-estimation}, in machine learning they are indicative of the topology of the loss landscape~\cite{ghorbani-2019-investigation-neural, yao-2020-pyhessian-neural}, and in graph processing they can uncover hidden graph motifs~\cite{huang-2021-density-states}. Given a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$, the distribution of its eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$, which places a rescaled Dirac delta distribution $\delta$ at each eigenvalue. Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation that is often prohibitively expensive. Several techniques have been developed for approximating $\phi$, including moment matching~\cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time} and polynomial approximation, either implicitly by a Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or explicitly by expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}. Some of these methods have also been combined with a deflation step to yield even more efficient algorithms \cite{lin-2017-randomized-estimation, bhattacharjee-2024-improved-spectral}.

@@

\paragraph{New contributions.} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices with constant randomness (with respect to $t$): The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}. Combined with Chebyshev approximation, this allows one to reuse the majority of computations across different values of the parameter $t$, making the estimators scale favorably with respect to the number of parameter evaluations. We propose several additional modifications to the methods from \cite{lin-2017-randomized-estimation} for the special case of approximating the smoothed spectral density $\phi_{\sigma}$ from~\refequ{equ:smoothed-spectral-density}. In particular, we use a more rigorous approach that preserves non-negativity for approximating matrix functions in terms of Chebyshev polynomials. The resulting Chebyshev-Nyström++ estimator 
unifies all methods from~\cite{lin-2017-randomized-estimation} in a single algorithm, with the desirable side effect of \textcolor{red}{speeding up the \enquote{spectrum sweeping} method by orders of magnitude. ??? Where do we actually demonstrate this? Illustrate/spell this out more explicitly in the numerical experiments ???} Combining our analysis of the randomized trace estimators with standard error bounds for Chebyshev approximation leads to Theorem~\refthm{thm:chebyshev-nystrom}, which shows that $\mathcal O(\varepsilon^{-1})$ matrix-vector products suffice to guarantee an error $\varepsilon$.

>>>>

Spectral distributions of matrices and linear operators reveal important properties in numerous applications across physics, chemistry, engineering, and data science: For example, in electronic structure calculations, eigenvalues represent the energy levels that electrons occupy~\cite{drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure, lin-2017-randomized-estimation}, in machine learning they are indicative of the topology of the loss landscape~\cite{ghorbani-2019-investigation-neural, yao-2020-pyhessian-neural}, and in graph processing they can uncover hidden graph motifs~\cite{huang-2021-density-states}. Given a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$, the distribution of its eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$, which places a rescaled Dirac delta distribution $\delta$ at each eigenvalue. Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation that is often prohibitively expensive. Several techniques have been developed for approximating $\phi$, including moment matching~\cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time} and polynomial approximation, either implicitly by a Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or explicitly by expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}. Some of these methods have also been combined with a deflation step to yield even more efficient algorithms \cite{lin-2017-randomized-estimation, bhattacharjee-2025-improved-spectral}.

@@

\paragraph{New contributions.} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices with constant randomness (with respect to $t$): The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}. Combined with Chebyshev approximation, this allows one to reuse the majority of computations across different values of the parameter $t$, making the estimators scale favorably with respect to the number of parameter evaluations. We propose several additional modifications to the methods from \cite{lin-2017-randomized-estimation} for the special case of approximating the smoothed spectral density $\phi_{\sigma}$ from~\refequ{equ:smoothed-spectral-density}. In particular, we use a more rigorous approach that preserves non-negativity for approximating matrix functions in terms of Chebyshev polynomials. The resulting Chebyshev-Nyström++ estimator 
unifies all methods from~\cite{lin-2017-randomized-estimation} in a single algorithm. %, and improves multiple algorithmic aspects to make them faster and more robust. %\textcolor{red}{speeding up the \enquote{spectrum sweeping} method by orders of magnitude. ??? Where do we actually demonstrate this? Illustrate/spell this out more explicitly in the numerical experiments ???}
Combining our analysis of the randomized trace estimators with standard error bounds for Chebyshev approximation leads to~\refthm{thm:chebyshev-nystrom}, which shows that $\mathcal O(\varepsilon^{-1})$ matrix-vector products suffice to guarantee an error $\varepsilon$.

### 2025-02-25 paper/numresults.tex
address review comments, clean up equations, create n_c=2 plot

While theoretically an important tool, we have observed that the preservation of non-negativity discussed in Section~\ref{sec:preservnonneg} is not needed in practice. The (slight) indefiniteness of the standard Chebyshev approximation $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ from~\refequ{equ:matrix-approximation} does not seem to impede the reliability of the Nyström++ estimator. Unless otherwise stated, we transform the input matrices such that their eigenvalues are contained in $[-1, 1]$, by approximating their spectrum with NumPy's \textcolor{red}{Hessian ??? replace by sparse? ???} eigenvalue solver. We compute spectral densities at $n_t = 100$ uniformly distributed values of the parameter $t \in [-1, 1]$. The integral involved in computing the $L^1$-errors is approximated with the composite midpoint quadrature rule using the $n_t$ values of $t$ as nodes. Both parameters introduced at the end \refsec{subsubsec:chebyshev-nystrom-implementation},the eigenvalue truncation threshold and the parameter for detecting a vanishing spectral density, are fixed to $10^{-5}$.

@@

We use the electronic structure interaction example from~\cite[Section 6]{lin-2017-randomized-estimation}, which involves the second order finite difference discretization of the Hamiltonian
\begin{equation}
    \mathcal{H} = - \Delta + V
    \label{equ:5-experiments-electronic-hamiltonian}
\end{equation}
in three dimensions. The potential $V$ interacting with the electrons is generated by Gaussian wells
%\begin{equation}
    $v(r) = v_0 e^{-\lambda r^2}$, 
%    \label{equ:5-experiments-gaussian-cell}
%\end{equation}
with $v_0 = -4$ and $\lambda = 8$, centered in cells of side length $L=6$ which are stacked $n_c \in \mathbb{N}$ times in each spatial dimension; see \reffig{fig:gaussian-well}. The grid width of the finite difference discretization is fixed to $h=0.6$. For example, for $n_c = 1$, this leads to a matrix of size $1000 \times 1000$. This example represents an idealized model for the interaction of nuclei on a regular grid with electrons for a $k$-vector in the center of the first Brillouin zone. The distribution of the eigenvalues of the Hamiltonian --- its spectral density --- allows one to interpret the system's energy levels.

\textcolor{red}{??? This discussion of $n_c$ only makes sense if we have at least one experiment with $n_c = 5$ or so. Could you please include such an experiment? For example, do Figure 4.2 again for larger $n_c$? Just use an overapproximation (large $n_\Psi$, $n_\Omega$) for approximating the true spectral distribution. ???}

@@

\reffig{fig:convergence} displays the convergence behavior of Chebyshev-Nyström++ applied to the $1000 \times 1000$ matrix resulting from the discretization of the Hamiltonian for $n_c = 1$. As a reference, we use the eigenvalues computed by NumPy's dense eigenvalue solver.
\begin{figure}[ht]
    \centering
    \input{plots/convergence.pgf}
    \caption{Error vs. number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ (with fixed $\sigma=0.005$ and \textcolor{red}{$m$ ??? Provide value???}) when applying  Chebyshev-Nyström++ to the Hamiltonian matrix from~\refsec{subsec:hamiltonian} with $n_c = 1$.}
    \label{fig:convergence}
\end{figure}%
As expected from \refthm{thm:hutchinson}, Girard-Hutchinson alone ($n_{\mtx{\Omega}} = 0$) require $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ samples to achieve an error of order $\varepsilon$. Because the eigenvalues of the matrix are spread out, we observe the better convergence of the Nyström approximation discussed at the end of \refsec{sec:application} once $n_{\mtx{\Omega}}$ is sufficiently large. \textcolor{red}{??? Comment explicitly that and why the errors levels off at around $10^{-6}$. Is it because of Chebyshev or some other truncation, etc.? ???}

@@

\subsection{Comparison to Krylov-aware stochastic trace estimation}
\label{subsec:krylov-aware}

The Krylov-aware stochastic trace estimator from \cite[Algorithm 3.1]{chen-2023-krylovaware-stochastic} is a viable alternative method for spectral density estimation. \textcolor{red}{??? Please rewrite this. It does not come out whether~\cite{chen-2023-krylovaware-stochastic} actually cover spectral density estimation. Are the Gaussian matrices chosen constants in their paper/in our experiments? If this is not discussed in~\cite{chen-2023-krylovaware-stochastic}, comment more explicitly to which extent this algorithm benefits from constant random matrices. Do they have an analysis that applies to our situation, etc. ??? It also samples two Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$. $\mtx{\Omega}$ is used to extract a low-rank approximation to $g_{\sigma}(t \mtx{I}_n - \mtx{A})$ by running the block-Lanczos algorithm on $\mtx{A}$ with starting block $\mtx{\Omega}$ for $r$ iterations with reorthogonalization and subsequently $q$ without. The columns of $\mtx{\Psi}$ are then used to estimate the trace of the approximation residual with $r$ iterations of the Lanczos algorithm on $\mtx{A}$ ??? Why $\mtx{A}$? ???.}

@@

When fitting the parameters of a neural network, it can be of interest to determine whether a (local) minimum has been found and whether this minimum is robust, i.e., whether a small perturbation of the parameters leads to a significant decrease in the fit of the neural network. Both properties are reflected in the eigenvalues of the Hessian matrix $\mtx{A}$ with respect to some loss function: if all eigenvalues are non-negative, then the loss attains a local minimum, and if, additionally, all of them are small, the minimum is robust. \textcolor{red}{??? Isn't it the opposite? If the eigenvalues are not close to zero, we get a robust local minimum? Please check! ???}

@@

We plot the approximation of the spectral density of the Hessian matrix of the untrained neural network, and at different stages of training in \reffig{fig:hessian-density}, as well as the corresponding mean squared error loss. It can be observed that the eigenvalues creep towards zero as the training proceeds. Furthermore, despite the loss steadily decreasing, the presence of relatively large eigenvalues in some epochs indicates a sharp minimum of the network, hence, unfavorable generalization properties.

@@

To demonstrate that the Chebyshev-Nyström++ estimator is also effective for problems unrelated to spectral density estimation, we consider the problem of approximating the partition function $\Trace(\exp(-\beta \mtx{A}))$ for the Hamiltonian $\mtx{A}$ of a planar Heisenberg spin chain from~\cite[Section 5.1]{chen-2023-krylovaware-stochastic}. We compute the pointwise error from the analytic solution as a function of the temperature $t\equiv \beta^{-1}$ and compare it to the error achieved by the Krylov-aware trace estimator in \reffig{fig:krylov-aware-spin}.

@@

Unlike for the spectral density example from \refsec{subsec:hamiltonian}, where the numerical rank of the matrix is relatively uniform across different $t$, this now changes changes wildly with the parameter $\beta$. Recall that the Chebyshev-Nyström++ method requires a fixed allocation of random vectors to the low-rank approximation for all values of the parameter $\beta$. Hence, both for large $\beta$, where the matrix $\exp(-\beta \mtx{A})$ has low numerical rank, and for small $\beta$, where the singular values decay very slowly, the approximation error is relatively large.
\textcolor{red}{??? Could you please recheck this example? Why is CN++ extremely bad for small $\beta^{-1}$? and KA much better? Please comment on this aspect explicitly. ???}

>>>>

While theoretically an important tool, we have observed that the preservation of non-negativity discussed in~\refsec{sec:preservnonneg} is not needed in practice. The (slight) indefiniteness of the standard Chebyshev approximation $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ from~\refequ{equ:matrix-approximation} does not seem to impede the reliability of the Nyström++ estimator. Unless otherwise stated, we transform the input matrices such that their eigenvalues are contained in $[-1, 1]$, by approximating their spectrum with NumPy's Hermitian eigenvalue solver before we apply the methods to them. We compute spectral densities at $n_t = 100$ uniformly distributed values of the parameter $t \in [-1, 1]$. The integral involved in computing the $L^1$-errors is approximated with the composite midpoint quadrature rule using the $n_t$ values of $t$ as nodes. Both parameters introduced at the end \refsec{subsubsec:chebyshev-nystrom-implementation},the eigenvalue truncation threshold and the parameter for detecting a vanishing spectral density, are fixed to $10^{-5}$.

@@

We use the electronic structure interaction example from~\cite[Section 6]{lin-2017-randomized-estimation}, which involves the second order finite difference discretization of the Hamiltonian
\begin{equation}
    \mathcal{H} = - \Delta + V
    \label{equ:electronic-hamiltonian}
\end{equation}
in three dimensions. The potential $V$ interacting with the electrons is generated by Gaussian wells
%\begin{equation}
    $v(r) = v_0 e^{-\lambda r^2}$, 
%    \label{equ:5-experiments-gaussian-cell}
%\end{equation}
with $v_0 = -4$ and $\lambda = 8$, centered in cells of side length $L=6$ which are stacked $n_c \in \mathbb{N}$ times in each spatial dimension; see \reffig{fig:gaussian-well}. The grid width of the finite difference discretization is fixed to $h=0.6$. For $n_c = 1$, this leads to a sparse matrix of size $1\,000 \times 1\,000$ and for $n_c = 2$ of size $8\,000 \times 8\,000$. This example represents an idealized model for the interaction of nuclei on a regular grid with electrons for a $k$-vector in the center of the first Brillouin zone. The distribution of the eigenvalues of the Hamiltonian --- its spectral density --- allows one to interpret the system's energy levels.

%\textcolor{red}{??? This discussion of $n_c$ only makes sense if we have at least one experiment with $n_c = 5$ or so. Could you please include such an experiment? For example, do Figure 4.2 again for larger $n_c$? Just use an overapproximation (large $n_\Psi$, $n_\Omega$) for approximating the true spectral distribution. ???}

@@

\reffig{fig:convergence} displays the convergence behavior of Chebyshev-Nyström++ applied to the matrix resulting from the discretization of the Hamiltonian for $n_c = 1$ and $n_c = 2$. As a reference, we use the eigenvalues computed by NumPy's dense eigenvalue solver.
\begin{figure}[ht]
    \centering
    \begin{subfigure}[b]{0.495\textwidth}
        \input{plots/convergence-1.pgf}
        \caption{$n_c = 1$}
        \label{fig:convergence-1}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \input{plots/convergence-2.pgf}
        \caption{$n_c = 2$}
        \label{fig:convergence-2}
    \end{subfigure}
    \caption{Error vs. number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ (with fixed $\sigma=0.005$ and $m=2000$) when applying  Chebyshev-Nyström++ to the Hamiltonian matrix from~\refsec{subsec:hamiltonian}.}
    \label{fig:convergence}
\end{figure}%
%\begin{figure}[ht]
%    \centering
%    \input{plots/convergence.pgf}
%    \caption{Error vs. number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ (with fixed $\sigma=0.005$ and $m=2000$) when applying  Chebyshev-Nyström++ to the Hamiltonian matrix from~\refsec{subsec:hamiltonian} with $n_c = 1$.}
%    \label{fig:convergence}
%\end{figure}%
As expected from \refthm{thm:hutchinson}, Girard-Hutchinson alone ($n_{\mtx{\Omega}} = 0$) require $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ samples to achieve an error of order $\varepsilon$. Because the eigenvalues of the matrix are spread out, we observe the better convergence of the Nyström approximation discussed at the end of \refsec{sec:application} once $n_{\mtx{\Omega}}$ is sufficiently large. We can observe the errors stagnate at around $10^{-6}$ as a consequence of the error made in the Chebyshev approximation.

%\begin{figure}[ht]
%    \centering
%    \input{plots/convergence_8_sigma005.pgf}
%    \caption{Error vs. number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ (with fixed $\sigma=0.005$ and $m=2000$) when applying  Chebyshev-Nyström++ to the Hamiltonian matrix from~\refsec{subsec:hamiltonian} with $n_c = 2$.}
%    \label{fig:convergence}
%\end{figure}%

@@

\subsection{Comparison to Krylov-aware trace estimator}
\label{subsec:krylov-aware}

We notice that the Krylov-aware stochastic trace estimator from \cite[Algorithm 3.1]{chen-2023-krylovaware-stochastic} can also effectively be used in the setting of spectral density approximation. It also samples two Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$. $\mtx{\Omega}$. It first runs block Lanczos algorithm on $\mtx{A}$ with starting block $\mtx{\Omega}$ for $r$ iterations with reorthogonalization and subsequently $q$ without. The columns of $\mtx{\Psi}$ are then projected onto the complement of the block Krylov subspace generated by the Lanczos algorithm, and are afterwards used as starting vectors for $r$ iterations of the Lanczos algorithm on $\mtx{A}$. What is remarkable is that all these heavy computations are completely independent of the smoothing kernel $g_{\sigma}$ and the parameter $t$. Further, some algebraic manipulations show that similarly to the stochastic Lanczos quadrature \cite[Section 3]{ubaru-2017-fast-estimation}, we can compress the estimator into a quadrature. Since this quadrature is independent of $g_{\sigma}$ and $t$, it can be applied to different smoothing kernels $g_{\sigma}$ and to as many values of $t$ as needed --- at little additional cost.

%\textcolor{red}{??? Please rewrite this. It does not come out whether~\cite{chen-2023-krylovaware-stochastic} actually cover spectral density estimation. Are the Gaussian matrices chosen constants in their paper/in our experiments? If this is not discussed in~\cite{chen-2023-krylovaware-stochastic}, comment more explicitly to which extent this algorithm benefits from constant random matrices. Do they have an analysis that applies to our situation, etc. ??? It also samples two Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$. $\mtx{\Omega}$ is used to extract a low-rank approximation to $g_{\sigma}(t \mtx{I}_n - \mtx{A})$ by running the block-Lanczos algorithm on $\mtx{A}$ with starting block $\mtx{\Omega}$ for $r$ iterations with reorthogonalization and subsequently $q$ without. The columns of $\mtx{\Psi}$ are then used to estimate the trace of the approximation residual with $r$ iterations of the Lanczos algorithm on $\mtx{A}$ ??? Why $\mtx{A}$? ???.}

@@

When fitting the parameters of a neural network, it can be of interest to determine whether a (local) minimum has been found and whether this minimum is robust, i.e., whether a small perturbation of the parameters leads to a significant decrease in the fit of the neural network. Both properties are reflected in the eigenvalues of the Hessian matrix $\mtx{A}$ with respect to some loss function: if all eigenvalues are non-negative, then the loss attains a local minimum, and if, additionally, all of them are small, we talk about a flat minimum; one where small perturbations of the parameters do not cause large increases of the loss and are therefore associated with better generalization \cite{yao-2020-pyhessian-neural}. On the other hand, large eigenvalues in the spectrum indicate high sensitivity of the loss to small changes in the parameters. Those we want to avoid, which is the reason why we monitor them.

@@

We plot the spectral density of the Hessian matrix of the untrained neural network, and at different stages of training in \reffig{fig:hessian-density}, as well as the corresponding mean squared error loss. It can be observed that the eigenvalues creep towards zero as the training proceeds. Furthermore, despite the loss steadily decreasing, the presence of relatively large eigenvalues in some epochs indicates a sharp minimum of the network, hence, unfavorable generalization properties.

@@

To demonstrate that the Chebyshev-Nyström++ estimator is also effective for problems unrelated to spectral density estimation, we consider the problem of approximating the partition function $\Trace(\exp(-\beta \mtx{A}))$ for the $1\,048\,576 \times 1\,048\,576$ Hamiltonian $\mtx{A}$ of a planar Heisenberg spin chain from~\cite[Section 5.1]{chen-2023-krylovaware-stochastic}. We compute the pointwise error from the analytic solution as a function of the temperature $t\equiv \beta^{-1}$ and compare it to the error achieved by the Krylov-aware trace estimator in \reffig{fig:krylov-aware-spin}.

@@

Unlike for the spectral density example from \refsec{subsec:hamiltonian}, where the numerical rank of the matrix is relatively uniform across different $t$, this now changes changes wildly with the parameter $\beta$. Recall that the Chebyshev-Nyström++ method requires a fixed allocation of random vectors to the low-rank approximation for all values of the parameter $\beta$. Hence, both for large $\beta$, where the matrix $\exp(-\beta \mtx{A})$ has low numerical rank, and for small $\beta$, where the singular values decay very slowly, the approximation error is relatively large. Similarly, one degree $m$ for the Chebyshev approximation needs to be fixed for all $\beta$. Clearly, for large values $\beta$, where $s \mapsto \exp(-\beta s)$ becomes difficult to approximate with polynomials, the fixed degree $m$ is not sufficient for a good Chebyshev approximation, and consequently, the Chebyshev-Nyström++ method is very inaccurate.

### 2025-02-25 paper/paper.tex
address review comments, clean up equations, create n_c=2 plot

\title{Randomized \textcolor{red}{I would prefer Stochastic} trace estimation for parameter-dependent matrices applied to spectral density approximation}

@@

\begin{abstract}
    Stochastic trace estimation is a well-established tool for approximating the trace of a large symmetric positive semi-definite matrix $\mtx{B}$. Several applications involve a matrix that depends continuously on parameter $t \in [a,b]$, and require trace estimates of $\mtx{B}(t)$ for many different values of $t$. This is, for example, the case when approximating the spectral density of a matrix or computing partition functions of quantum systems. Approximating the trace separately for each matrix
    $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ clearly incurs redundancies and a cost that scales linearly with $m$. To address this issue, we propose and analyze modifications for three stochastic trace estimators, the Girard-Hutchinson, Nyström, and Nyström++ estimators. Our modification uses \emph{constant} randomization across different values of $t$, that is,
    every matrix $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ is multiplied with the \emph{same} set of random vectors.
    When combined with Chebyshev approximation in $t$, the use of such constant random matrices allows one to reuse matrix vector-products across different values of $t$, leading to significant cost reduction.
    Our analysis shows that the loss of stochastic independence across different $t$ does not lead to detoriation; in fact, our error bounds for these estimators closely match existing error bounds for the constant case. In particular, we show that $\mathcal{O}(\varepsilon^{-1})$ random matrix-vector products suffice to ensure an error of $\varepsilon > 0$ for Nyström++, independent of low-rank properties of $\mtx{B}(t)$. We discuss in detail how the combination of Nyström++ with 
    Chebyshev approximation applies to large-scale spectral density estimation and provide an analysis of the resulting method. This  improves various aspects of an existing stochastic estimator for spectral density estimation. Several numerical experiments for examples from electronic structure interaction, statistical thermodynamics, and neural network optimization illustrate and validate our findings.
\end{abstract}

\textcolor{red}{TODO for Fabio:
\begin{itemize}
 \item Has~\cite{bhattacharjee-2024-improved-spectral} been published?
 \item Fix discussion end of Section 3.3.2
 \item Please reread entire Section 4 and address various red points.
 \item Please reread abstract.
 \item Please reread last part of Section 2 (startin from new contributions) and fix red remark.
 \item Please remove all unreferenced equation labels (using checkref) at the end.
 \item Perform a spell check at the end.
\end{itemize}
}
\input{intro.tex}
\input{analysis.tex}
\input{spectraldensity.tex}
\input{numresults.tex}
\input{conclusion.tex}

>>>>

\title{Stochastic trace estimation for parameter-dependent matrices applied to spectral density approximation}

@@

\begin{abstract}
    Stochastic trace estimation is a well-established tool for approximating the trace of a large symmetric positive semi-definite matrix $\mtx{B}$. Several applications involve a matrix that depends continuously on a parameter $t \in [a,b]$, and require trace estimates of $\mtx{B}(t)$ for many different values of $t$. This is, for example, the case when approximating the spectral density of a matrix or computing partition functions of quantum systems. Approximating the trace separately for each matrix
    $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ clearly incurs redundancies and a cost that scales linearly with $m$. To address this issue, we propose and analyze modifications for three stochastic trace estimators, the Girard-Hutchinson, Nyström, and Nyström++ estimators. Our modification uses \emph{constant} randomization across different values of $t$, that is,
    every matrix $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ is multiplied with the \emph{same} set of random vectors.
    When combined with Chebyshev approximation in $t$, the use of such constant random matrices allows one to reuse matrix-vector products across different values of $t$, leading to significant cost reduction.
    Our analysis shows that the loss of stochastic independence across different $t$ does not lead to deterioriation; in fact, our error bounds for these estimators closely match existing error bounds for the constant case. In particular, we show that $\mathcal{O}(\varepsilon^{-1})$ random matrix-vector products suffice to ensure an error of $\varepsilon > 0$ for Nyström++, independent of low-rank properties of $\mtx{B}(t)$. We discuss in detail how the combination of Nyström++ with 
    Chebyshev approximation applies to large-scale spectral density estimation and provide an analysis of the resulting method. This  improves various aspects of an existing stochastic estimator for spectral density estimation. Several numerical experiments for examples from electronic structure interaction, statistical thermodynamics, and neural network optimization illustrate and validate our findings.
\end{abstract}

%\textcolor{red}{TODO for Fabio:
%\begin{itemize}
 %\item Has~\cite{bhattacharjee-2024-improved-spectral} been published?
 %\item Fix discussion end of Section 3.3.2
 %\item Please reread entire Section 4 and address various red points.
 %\item Please reread abstract.
 %\item Please reread last part of Section 2 (startin from new contributions) and fix red remark.
 %\item Please remove all unreferenced equation labels (using checkref) at the end.
 %\item Perform a spell check at the end.
%\end{itemize}
%}
\input{intro.tex}
\input{analysis.tex}
\input{spectraldensity.tex}
\input{numresults.tex}
\input{conclusion.tex}
